Staged Abstract Interpreters

### 2018-11-17 paper/10_intro.tex
paper
revise

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close connection between 
compilers and interpreters. 
The first Futamura projection specifically shows that specializing an interpreter with respect to 
the input program yields an equivalent executable.
Partial evaluation \cite{DBLP:books/daglib/0072559} was the first proposed approach to realize 
Futamura projections, which first identifies the binding-time of variables in the program, i.e., 
they can be known whether statically or dynamically, and then evaluates the static part, and 
generates a residual program that relies on the dynamic part.
However, precisely analyzing binding-time for an arbitrary program is hard in general. As an 
alternative and programmatic approach to specialization and partial evaluation, multi-stage 
programming (MSP for shot) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the stage 
annotations (i.e., binding-time annotations) from the programmers. These staging annotations 
identify which part of the input program should be specialized.
As a classical example, we consider a power function: 
\begin{lstlisting}
def power(b: Int, e: Int): Int = 
  if (e == 0) 1 else e * power(b, e-1)
\end{lstlisting}
If \texttt{e} is identified as statically known by the programmer, say 5, then we can 
specialize the function \texttt{power} with respect to \texttt{e = 5} and generate a new 
specialized function that runs faster:
\begin{lstlisting}
def power5(b: Int) = b * b * b * b * b
\end{lstlisting}

Abstract interpretation \cite{DBLP:conf/popl/CousotC77} as a semantic approach to construct 
sound static analysis by approximation is almost as old as Futamura projection.
Some recent progress such as Abstracting Abstract Machines (AAM) uncovers a methodology to 
derive sound abstract interpreters from their concrete counterparts and has been applied to 
different variants of definitional interpreters and abstract machines
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}.
Given the structural similarity between concrete and abstract interpreters, intellectually 
it is natural to raise the question whether it is possible to specialize an abstract interpreter, and how 
can we effectively do that having considered their functional dissimilarity. 
This paper studies the confluence of two old ideas --- Futamura project and abstract interpretation,
but from the perspective of their recent realizations --- multi-stage programming and definitional
abstract interpreters. To be specific, we present the application of the first Futamura projection
on abstract interpreters, and the approach of constructing optimizing abstract interpreters by 
multi-stage programming.

\begin{figure}[h]
  \begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em]
  {
    \begin{smallmatrix} \text{unstaged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} \\
      \begin{smallmatrix} \text{unstaged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} \\ };
  \path[-stealth]
    (m-1-1) edge node [above, font=\scriptsize] {$\updownarrows$} (m-1-2)
    (m-2-1) edge node [left, font=\scriptsize] {$\alpha$} (m-1-1)
    (m-2-1) edge node [below, font=\scriptsize] {$\updownarrows$} (m-2-2)
    (m-2-2) edge node [right, font=\scriptsize] {$\alpha$} (m-1-2);
  \end{tikzpicture}
  \caption{The confluence of staging and abstraction.}
\end{figure}

Multi-stage programming has been widely used to improve the performance in many domains, 
such as optimizing compilers and domain-specific languages \cite{DBLP:conf/pldi/RompfSBLCO14, DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13, DBLP:journals/jfp/CaretteKS09},
numerical computation \cite{PGL-038, DBLP:conf/pepm/AktemurKKS13}, 
generic programming \cite{DBLP:journals/pacmpl/Yallop17, Ofenbeck:2017:SGP:3136040.3136060}, 
data processing \cite{DBLP:conf/oopsla/JonnalageddaCSRO14, DBLP:conf/popl/KiselyovBPS17}, 
query compilation in databases \cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18},
etc.
Likewise, specializing static analyses by partial evaluation emerged in late 90s 
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587, ashley:practical}, 
and indeed it is able to effectively remove the interpretive overhead of repeatedly traversing 
the abstract syntax tree. 
However, despite of the previous works mostly focused on one particular analysis, or required to 
completely rewrite the analyzer,
it is worth to investigate the idea from a modern perspective based on generative programming,
especially for a general abstract interpreter that models direct-style $\lambda$-calculus, imperative 
features such as mutation, and different abstract domains.
One technical challenge is the binding-time engineering when non-determinism, 
fixed-point iterations, different abstract domains and staging are introduced at the same time.
In this paper, we present an end-to-end design and implementation of staging an abstract interpreter,
that means not only the interpreter that traverses the abstract syntax tree, 
but also the data structure of abstract domains, abstract environment and heap, 
and the fixed-point iteration are staged.

@@

When targeting higher-order programs, either staged concrete interpreters or staged abstract interpreters
are able to compile a closure, i.e., specialize the call of interpreter with respect to the body
expression of the lambda term without knowing the actual argument value. By generalizing this observation,
we can actually specialize the abstract interpreter with any open programs, 
which unexpectedly leads to a modular analysis and improves scalability.
A open program contains free variables, which represent other parts of the program and will be left
as dynamic values.
For instances, one challenge in static analysis of modern software is that the programs are usually 
shipped with large library code \cite{toman_et_al:LIPIcs:2017:7121}, for example, it has been shown that 
analyzing a simple "Hello World" program in Java depends on additional 3,000 classes in the library 
\cite{DBLP:conf/oopsla/KulkarniMZN16}.
A precise whole-program analysis formulated as abstract interpreters can be very expensive
due to the scale of the program and the inherent complexity of the algorithm.
However, analyzing these libraries sometimes are unnecessarily repeated.
When applying the staged abstract interpreter on such open programs, we leave the unknown arguments and 
calling contexts as dynamic values and generate partial analyzing result which is 
represented as a residual program. The partial analyzing result can be reused and composed with the 
analyzing result of application code when available later. Therefore we mechanically obtain a modular
analysis through staging, even though the original algorithm is formulated as a whole-program analysis.

@@

\begin{itemize}
  \item Starting from a generic definition of interpreter for a higher-order functional language,
    whose syntax and semantics are described in Section~\ref{bg_lang}, we show its concrete
    interpreter (Section~\ref{unstaged_conc}), staged concrete interpreter (Section~\ref{stagedinterp}), 
    abstract interpreter (Section~\ref{unstaged_abs}) and staged abstract interpreter (Section~\ref{sai}).
    Our design allows one can derive these four different kinds of interpreters by mechanically changing
    the underlying semantics or stage annotations.
    %We later extend the staged abstract interpreter to support imperative language features (Section~\ref{cases_imp}), such as states and loops.
  \item We demonstrate that the staged abstract interpreter not only improves the \textit{efficiency}
    but also the \textit{scalability}, which are two major issues in static analysis, if we apply it on open
    programs. 
    %\todo{more on this point.}
  \item To demonstrate the applicability and usefulness of the proposed approach, we conduct three case studies 
    in Section~\ref{cases_study}. 
    \subitem (1) We revisit Abstract Compilation \cite{Boucher:1996:ACN:647473.727587} in Section~\ref{cs_ac}, 
    which was originally a closure generation technique applied on 0-CFA.
    We show that abstract compilation can be understood and easily implemented as an instance of our staged abstract interpreter.
    But unlike the abstract compilation, our approach does not require the implementers to rewrite the analysis.
    \subitem (2) We demonstrate pushdown control-flow analysis techniques (Section~\ref{cfa}), such as context-sensitive 
    (\todo{adaptive/flow})
    analysis and more precise stores can be integrated to staged abstract interpreters without losing any precision.
    \subitem (3) We show that in an imperative language (Section~\ref{cases_imp}), by replacing the abstract domain
    to an interval domain (as well as staged), the staged abstract interpreter performs numerical analysis.
    % (Optional: taint/information flow analysis?)
  \item We empirically evaluate the performance improvement by staging an big-step abstract interpreter 
    for control-flow analysis in Section~\ref{evaluation}. 
    We compare our staged abstract interpreter with the unstaged version and a small-step abstract interpreter.
\end{itemize}

>>>>

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close connection between 
compilers and interpreters. 
The first Futamura projection specifically shows that specializing an interpreter with respect to 
the input program yields an equivalent executable.
Partial evaluation \cite{DBLP:books/daglib/0072559} was the first proposed approach to realize 
Futamura projections, which first identifies the binding-time of variables in the program, i.e., 
they can be known whether statically or dynamically, and then evaluates the static part, and 
generates a residual program that relies on the dynamic part.
However, precisely analyzing binding-time for an arbitrary program is hard in general. As an 
alternative and pragmatic approach to specialization and partial evaluation, multi-stage 
programming (MSP for short) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the stage 
annotations (i.e., binding-time annotations) from the programmers. These staging annotations 
identify which part of the input program should be specialized.
As a classical example, we consider a power function: 
\begin{lstlisting}
def power(b: Rep[Int], e: Int): Int = 
  if (e == 0) 1 else e * power(b, e-1)
\end{lstlisting}
If the program identifies that @b@ will be known dynamically (as shown in its type @Rep[Int]@ -- a representation of @Int@) 
and @e@ is statically known, say 5, 
then we can specialize the function \texttt{power} with respect to \texttt{e = 5} and generate a new 
specialized function that runs faster:
\begin{lstlisting}
def power5(b: Int) = b * b * b * b * b
\end{lstlisting}

Abstract interpretation \cite{DBLP:conf/popl/CousotC77} as a semantic approach to construct 
sound static analysis by approximation is almost as old as Futamura projections.
Some recent progress such as Abstracting Abstract Machines (AAM) uncovers a methodology to 
derive sound abstract interpreters from their concrete counterparts and has been applied to 
different variants of definitional interpreters and abstract machines
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}.
Given the structural similarity between concrete and abstract interpreters, intellectually 
it is natural to raise the question whether it is possible to specialize an abstract interpreter, and how 
can we effectively do that having considered their different functionalities.
This paper studies the confluence of two old ideas --- Futamura project and abstract interpretation,
but from the perspective of their recent realizations --- multi-stage programming and definitional
abstract interpreters. To be specific, we present the application of the first Futamura projection
on abstract interpreters, and the approach of constructing optimizing abstract interpreters by 
multi-stage programming.

\begin{figure}[h]
  \begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em]
  {
    \begin{smallmatrix} \text{unstaged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} \\
      \begin{smallmatrix} \text{unstaged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} \\ };
  \path[-stealth]
    (m-1-1) edge node [above, font=\scriptsize] {$\updownarrows$} (m-1-2)
    (m-2-1) edge node [left, font=\scriptsize] {$\alpha$} (m-1-1)
    (m-2-1) edge node [below, font=\scriptsize] {$\updownarrows$} (m-2-2)
    (m-2-2) edge node [right, font=\scriptsize] {$\alpha$} (m-1-2);
  \end{tikzpicture}
  \caption{The confluence of staging and abstraction.}
  \label{confluence}
\end{figure}

Multi-stage programming has been widely used to improve the performance in many domains, 
such as optimizing compilers or domain-specific languages \cite{DBLP:conf/pldi/RompfSBLCO14, DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13, DBLP:journals/jfp/CaretteKS09},
numerical computation \cite{PGL-038, DBLP:conf/pepm/AktemurKKS13}, 
generic programming \cite{DBLP:journals/pacmpl/Yallop17, Ofenbeck:2017:SGP:3136040.3136060}, 
data processing \cite{DBLP:conf/oopsla/JonnalageddaCSRO14, DBLP:conf/popl/KiselyovBPS17}, 
query compilation in databases \cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18},
etc.
Likewise, specializing static analyses by partial evaluation emerged in late 90s 
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587, ashley:practical}, 
and indeed it is able to effectively remove the interpretive overhead of repeatedly traversing 
the abstract syntax tree. 
However, these previous works focused mostly on one particular analysis, or required to 
completely rewrite the analyzer. 
Hence, it is worth to investigate the idea from a modern perspective based on generative programming,
especially for a general abstract interpreter that models direct-style $\lambda$-calculus, imperative 
features such as mutation, and different abstract domains.
One technical challenge is the binding-time engineering when non-determinism, 
fixed-point iterations, different abstract domains and staging are introduced at the same time.
In this paper, we present an end-to-end design and implementation of staging an abstract interpreter;
that means not only the interpreter that traverses the abstract syntax tree, 
but also the data structure of abstract domains, abstract environment and heap, 
and the fixed-point iteration are staged.

@@

When targeting higher-order programs, either staged concrete interpreters or staged abstract interpreters
are able to compile a closure, i.e., specialize the call of interpreter with respect to the body
expression of the lambda term without knowing the actual argument value. By generalizing this observation,
we can actually specialize the abstract interpreter with any open programs, 
which unexpectedly leads to a modular analysis and improves scalability.
A open program contains free variables, which represent other parts of the program and will be left
as dynamic values.
For instances, one challenge in static analysis of modern software is that the programs are usually 
shipped with large library code \cite{toman_et_al:LIPIcs:2017:7121}, for example, it has been shown that 
analyzing a simple "Hello World" program in Java depends on additional 3,000 classes in the library 
\cite{DBLP:conf/oopsla/KulkarniMZN16}.
A precise whole-program analysis formulated as abstract interpreters can be very expensive
due to the scale of the program and the inherent complexity of the algorithm.
However, analyzing these libraries is sometimes unnecessarily repeated.
When applying the staged abstract interpreter on such open programs, we leave the unknown arguments and 
calling contexts as dynamic values and generate partial analyzing result which is 
represented as a residual program. The partial analyzing result can be reused and composed with the 
analyzing result of application code when available later. Therefore we mechanically obtain a modular
analysis through staging, even though the original algorithm is formulated as a whole-program analysis.

@@

This paper makes the following contributions:

\begin{itemize}
  \item Starting from a generic definition of interpreter for a higher-order functional language,
    whose syntax and semantics are described in Section~\ref{bg_lang}, we show its concrete
    interpreter (Section~\ref{unstaged_conc}), staged concrete interpreter (Section~\ref{stagedinterp}), 
    abstract interpreter (Section~\ref{unstaged_abs}) and staged abstract interpreter (Section~\ref{sai}).
    Our design allows one can derive these four different kinds of interpreters by mechanically changing
    the underlying semantics or stage annotations.
    %We later extend the staged abstract interpreter to support imperative language features (Section~\ref{cases_imp}), such as states and loops.
  \item We demonstrate that the staged abstract interpreter not only improves the \textit{efficiency}
    but also the \textit{scalability}, which are two major issues in static analysis, if we apply it on open
    programs. \todo{Section X}.
  \item To demonstrate the applicability and usefulness of the proposed approach, we conduct three case studies 
    Section~\ref{cases_study}: 
    \subitem (1) We revisit Abstract Compilation \cite{Boucher:1996:ACN:647473.727587} in Section~\ref{cs_ac}, 
    which was originally a closure generation technique applied on 0-CFA.
    We show that abstract compilation can be understood and easily implemented as an instance of our staged abstract interpreter.
    But unlike the abstract compilation, our approach does not require the implementers to rewrite the analysis.
    \subitem (2) We demonstrate pushdown control-flow analysis techniques (Section~\ref{cfa}), such as context-sensitive 
    analysis and more precise stores can be integrated to staged abstract interpreters without losing any precision.
    \subitem (3) We show that in an imperative language (Section~\ref{cases_imp}), by replacing the abstract domain
    to an interval domain (as well as staged), the staged abstract interpreter performs numerical analysis.
    % (Optional: taint/information flow analysis?)
  \item We empirically evaluate the performance improvement by staging an big-step abstract interpreter 
    for control-flow analysis (Section~\ref{evaluation}). 
    We compare our staged abstract interpreter with the unstaged version.
\end{itemize}

### 2018-11-17 paper/20_bg.tex
revise

\section{Preliminary}

In this section, we first describe the abstract syntax of the language we will analyze, then present an 
generic interface of interpreters, and instantiate an unstaged concrete interpreter.
Then, starting from this unstaged concrete interpreter, we proceed in two directions:
by abstracting it we obtain an unstaged abstract interpreter (Section~\ref{unstaged_abs}); 
by staging it we obtain a staged concrete interpreter (Section~\ref{stagedinterp}).

>>>>

\section{Preliminaries}

In this section, we first describe the abstract syntax of the language we will analyze, then present a
generic interface for interpreters, and instantiate an unstaged concrete interpreter.
Then, starting from this unstaged concrete interpreter, we proceed in two directions:
by abstracting it we obtain an unstaged abstract interpreter (Section~\ref{unstaged_abs}); 
by staging it we obtain a staged concrete interpreter (Section~\ref{stagedinterp}).

### 2018-11-17 paper/21_bg_interp.tex
revise intro
revise

As higher-order functions had been widely introduced to mainstream programming languages (such as Python and Java \todo{cite}), 
in this paper, we consider a tiny higher-order functional language \TLang, which is based on direct-style $\lambda$-calculus
with numbers, arithmetics, recursions and conditionals.
In Section~\ref{cases_imp}, we will see adding more imperative features to the language.
Since we are mostly interested in analyzing the dynamic behavior of the program, the static semantics and
type system will be omitted. We also assume the input programs are well-typed and variables are distinct.
The abstract syntax is described below:

@@

We give the concrete semantics of \TLang by showing a big-step operational-style interpreter for it.
The interpreter is a recursive function that takes the program AST, environment and store, and returns 
the evaluated value and the accompanied store. The environment will be a mapping from identifiers to addresses,
and the store is a mapping from addresses to values. The store simulates the heap and is useful when modeling 
recursion and mutation.
This environment-and-store-passing style big-step interpreter also can be obtained by 
refunctionalizing \cite{DBLP:conf/ppdp/AgerBDM03, Wei:2018:RAA:3243631.3236800} a small-step CESK 
machine \cite{DBLP:conf/popl/FelleisenF87}.

\subsection{Generic Interface} 
Before showing the actual implementation of the concrete semantics, let's first see 
a generic interface for the interpreters, which is semantics-agnostic and can be 
"interpreted" by either concrete or abstract, unstaged or staged semantics by our choice.
The trait \texttt{Semantics} first declares several abstract type members, such as \texttt{Addr},
\texttt{Value}, \texttt{Env}, \texttt{Store} and returned type \texttt{Ans}. 
Initial values for environments and stores, and few abstract methods that manipulate environments and stores are also declared.

@@

The actual semantics and operations of the interpreter is left to be implemented later, 
such as lifting literal terms to values (\texttt{close}, \texttt{num}), 
applying a closure (\texttt{apply\_closure}), branching (\texttt{branch0}),
and arithmetics (\texttt{prim\_eval}). We can choose to implement them concrete or abstract,
unstaged or staged.
But, just based on these abstract operations, an expressive generic interpreter can be defined:

@@

The implementation of the skeleton interpreter traverses the abstract syntax tree and should be straightforward.
It is worth noting that the interpreter is written in open-recursive style -- 
it can not refer to itself directly, instead, \texttt{eval} takes an additional 
parameter \texttt{ev} referring to itself; this allows us to instrument the interpreter
from outside. Accordingly, the function \texttt{close} for lifting lambda term to closure values
is also written in this style.
We define a default implementation \texttt{fix} to close the recursion by recursively using the result 
of \texttt{fix(ev)} for \texttt{ev}. 
Finally, the top-level evaluation function \texttt{eval\_top} is defined.

\paragraph{Stage Polymorphism.}
Readers may notice an abstract higher-kinded type \texttt{R[+T]} is defined in the trait, that is a
type-level function and will be used to specify the binding-time, i.e., staged or not; the annotation 
\texttt{+} indicates type parameter \texttt{T} is covariant.
The LMS framework we use here allows us to specify binding-time purely based on types.
All occurrences of type \texttt{Env}, \texttt{Store}, etc. are wrapped with \texttt{R} 
essentially because they may potentially be dynamic when staging is introduced.
Obviously, the unstaged higher-kinded type then can be simply instantiated as an identity type function: 
\texttt{type R[+T] = T}. In Section~\ref{stagedinterp}, \texttt{R} will be instantiated 
as a staging annotation. \todo{where talk about stage poly?}

\subsection{Concrete Instantiation}
Now we can instantiate a concrete interpreter. We first concretize the abstract types as follows.
Since the interpreter is concrete, for the propose that properly models the heap, the type \texttt{Int} 
is used for address space \texttt{Addr}. The environment and store are just ordinary \texttt{Map} in Scala.
The values can be either a tagged number \texttt{NumV} or a closure \texttt{CloV} that contains a lambda term
and an environment.

@@

The we implement the components that define the semantics. 
As we mentioned before, the type \texttt{R} is concretize as an identity type function, thus
all types are just ordinary types.

@@

\texttt{get} and \texttt{put} directly manipulate the map of environment and store. 
An allocation function is used when we need to update the store: 
in the concrete case, \texttt{alloc} always returns a fresh address of the store.
\texttt{close} and \texttt{num} lift syntactic literals to values.
\texttt{apply\_closure} takes the function value, argument value and the latest store,
then extends the environment and store after allocating a fresh address, and continue
evaluating the body \texttt{e} of the lambda term.

>>>>

We consider a tiny higher-order functional language \TLang, which is based on direct-style $\lambda$-calculus
with numbers, arithmetics, recursions and conditionals.
In Section~\ref{cases_imp}, we will add more imperative features to the language.
Since we are mostly interested in analyzing the dynamic behavior of the program, we disguised any static semantics and
type system. We also assume that input programs are well-typed and variables are distinct.
The abstract syntax is as follows:

@@

We give the concrete semantics of \TLang by showing a big-step definitional interpreter.
The interpreter is a recursive function that takes the program AST, environment and store, and returns 
the evaluated value and the accompanied store. The environment is a mapping from identifiers to addresses,
and the store is a mapping from addresses to values. 
We use the store to model recursion and mutation in concrete semantics; it is also useful for polyvariant analysis.
This environment-and-store-passing style big-step interpreter is standard and can also be obtained by 
refunctionalizing \cite{DBLP:conf/ppdp/AgerBDM03, Wei:2018:RAA:3243631.3236800} a small-step CESK 
machine \cite{DBLP:conf/popl/FelleisenF87}.

\subsection{Generic Interface} 
Before showing the actual implementation of the concrete semantics, let us first consider
a generic interface for the interpreters, which is semantics-agnostic and can be 
"interpreted" by either concrete or abstract, unstaged or staged semantics of our choice.
The trait \texttt{Semantics} first declares several abstract type members, such as \texttt{Addr},
\texttt{Value}, \texttt{Env}, \texttt{Store} and returned type \texttt{Ans}. 
It also declares initial values for environments and stores, and few abstract methods that 
manipulate environments and stores such as @put@ and @get@.

@@

The actual semantics and operations of the interpreter are left to be implemented later, 
these cinclude operations such as lifting literal terms to values (\texttt{close}, \texttt{num}), 
applying a closure (\texttt{apply\_closure}), branching (\texttt{branch0}),
and arithmetics (\texttt{prim\_eval}). We can choose to implement them in a way that is concrete or abstract,
unstaged or staged.
But, just based on these abstract operations, an expressive generic interpreter can be defined:

@@

The implementation of the skeleton interpreter traverses the abstract syntax tree and should be straightforward to understand.
It is worth noting that the interpreter is written in open-recursive style -- 
it can not refer to itself directly, instead, \texttt{eval} takes an additional 
parameter \texttt{ev} referring to itself; this allows us to instrument the interpreter
from the outside. Accordingly, the function \texttt{close} for lifting lambda term to closure values
is also written in this style.
We define a default implementation \texttt{fix} to close the recursion by recursively using the result 
of \texttt{fix(ev)} for \texttt{ev}. 
Finally, the top-level evaluation function \texttt{eval\_top} is defined.

\paragraph{Stage Polymorphism.}
Readers may notice that an abstract higher-kinded type \texttt{R[+T]} is defined in the trait, i.e. a
type-level function whici will be used to specify the binding-time, i.e., staged or not; the annotation 
\texttt{+} indicates that the type parameter \texttt{T} is covariant.
The LMS framework we use here allows us to specify binding-times purely based on types.
All occurrences of type \texttt{Env}, \texttt{Store}, etc. are wrapped with \texttt{R} 
essentially because they may potentially be dynamic when staging is introduced.
Obviously, the unstaged higher-kinded type then can be simply instantiated as an identity type function: 
\texttt{type R[+T] = T}. In Section~\ref{stagedinterp}, \texttt{R} will be instantiated 
as a staging annotation. \todo{Finally tagless, poly embedding}

\subsection{Concrete Instantiation} \label{unstaged_conc}
Now we can instantiate a concrete interpreter. We first concretize the abstract types as follows.
Since the interpreter is concrete, for the propose that properly models the heap, the type \texttt{Int} 
is used for address space \texttt{Addr}. The environment and store are just ordinary \texttt{Map}s in Scala.
A value can be either a tagged number \texttt{NumV} or a closure \texttt{CloV} that contains a lambda term
and an environment.

@@

we proceed to implement the components that define the semantics. 
As mentioned before, type \texttt{R} is concretized as an identity type function, thus
all types are just ordinary types.

@@

Finally, \texttt{get} and \texttt{put} directly manipulate the map of environment and store. 
An allocation function is used when we need to update the store: 
in the concrete case, \texttt{alloc} always returns a fresh address of the store.
\texttt{close} and \texttt{num} lift syntactic literals to values.
\texttt{apply\_closure} takes the function value, argument value and the latest store,
then extends the environment and store after allocating a fresh address, and continue
evaluating the body \texttt{e} of the lambda term.
